COVID19 - Forecast and predictions using a time dependent SEIR model for the Veneto Region

### 2020-03-24 draft_analysis_Italy_html.Rmd
Add home page, navigation and theme

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days by means of a linear regressione model, assuming a Normal distribution for the beta (the slope).
```{r R0 forecast}
time<-3:(days-2)
weekend<-rep(c(1,2,3,4,5,6,7),ceiling(days/7))[3:(days-2)]
data=data.frame(time,weekend)
beta.model<-glm(beta_vec~time+bs(weekend),weights = 1/sd_vec,family=gaussian,data=data)
summary(beta.model)
anova(beta.model,test="Chisq")
#there is an effect of the week, as supposed 
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
weekend_pre<-rep(c(1,2,3,4,5,6,7),ceiling((days+forecast)/7))[1:(days+forecast)]
datanew<-data.frame(time=1:(days+forecast),weekend=weekend_pre)
pre<-predict(beta.model,type='response',newdata=datanew,se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
predict <- data.frame(beta_vec=c(rep(NA,2),beta_vec,rep(NA,forecast+2)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)
beta.predict<-predict 
r0.predict<-beta.predict
r0.predict[,c(1,3:5)]<-r0.predict[,c(1,3:5)]*duration+1
# plot the points (actual observations), regression line, and confidence interval
p <- ggplot(r0.predict, aes(date,beta_vec))
p <- p + geom_point() +labs(x="Date",y="R0 value") 
p <- p + geom_line(aes(date,fit))
p <- p + geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)
p
```

@@

We made a forecast by means of a SEIR model fixing a series of initial status:
*S:N, the size of Italian population  
*E: The number of exposed people (unknown)  
*I_start: initial number of COVID-19 cases     
*R_start: initial number of recovered

@@

The parameter $\sigma$ and the fraction of exposed people ($E_{start}$) are our parameters of interest; we estimate $\sigma$ and $E_{start}$ minimizing the log likelihood of the described Poisson distribution (or Negative Binomial one).

@@

f1<-function(par){
f_exp<-expit(par[1])
sigma<-expit(par[2])
parameters <- c(mu = mu0, beta =(mean(beta_vec[1:(days-last-2)])+1/duration), sigma = sigma, gamma =1/duration)
initials <- c(S = (1-(f_exp+I_1/N-R_start/N)), E = f_exp, I = I_1/N, R = R_1/N)
-LogLikfun(initials,parameters,dat_csv$totale_attualmente_positivi[c((days-last):days)])
}
est<-optim(fn=f1,par=c(logit(0.0005),logit(0.1)))
#estimated par
expit(est$par)
f_exp<-expit(est$par[1])
sigma<-expit(est$par[2])
parameters <- c(mu = mu0, beta =(mean(beta_vec[1:(days-last-2)])+1/duration), sigma = sigma, gamma = 1/duration)
initials <- c(S = (1-(f_exp+I_1/N-R_1/N)), E = f_exp, I = I_1/N, R = R_1/N)
pro<-SEIR(pars = parameters, init = initials, time = 0:last)
#predicted vs observed
cbind(pro$results$I*N,dat_csv$totale_attualmente_positivi[c((days-last):days)])
```

@@

```{r status, echo=TRUE}
#mthe unknowrn fraction of exposed people is
pro$results$E[last+1];
#that is 
pro$results$E[last+1]/(I_1/N)
#times the infected
```
For the beta parameter, we perfom a simulation on its trend by means of a Bayesian Structural Time Series using the library bsts of R.    
We estimate a BSTS model specifing a local linear trend ans 1000 simulation.

>>>>

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days by means of a linear regressione model, assuming a Normal distribution for the beta (the slope).
```{r R0 forecast}
time<-3:(days-2)
weekend<-rep(c(1,2,3,4,5,6,7),ceiling(days/7))[3:(days-2)]
data=data.frame(time,weekend)
beta.model<-glm(beta_vec~time+I(cos(time*2*pi/7))+I(sin(time*2*pi/7)),weights = 1/sd_vec,family=gaussian,data=data)
summary(beta.model)
anova(beta.model,test="Chisq")
#there is an effect of the week, as supposed 
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
weekend_pre<-rep(c(1,2,3,4,5,6,7),ceiling((days+forecast)/7))[1:(days+forecast)]
datanew<-data.frame(time=1:(days+forecast),weekend=weekend_pre)
pre<-predict(beta.model,type='response',newdata=datanew,se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
predict <- data.frame(beta_vec=c(rep(NA,2),beta_vec,rep(NA,forecast+2)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)
beta.predict<-predict 
r0.predict<-beta.predict
r0.predict[,c(1,3:5)]<-r0.predict[,c(1,3:5)]*duration+1
# plot the points (actual observations), regression line, and confidence interval
p <- ggplot(r0.predict, aes(date,beta_vec))
p <- p + geom_point() +labs(x="Date",y="R0 value") 
p <- p + geom_line(aes(date,fit))
p <- p + geom_ribbon(aes(ymin=lwr,ymax=upr), alpha=0.3)
p
```

@@

We made a forecast by means of a SEIR model fixing a series of initial status:   
*S:N, the size of Italian population  
*E: The number of exposed people, but supposed to be at least 3 times the infected (3 x I_start)  
*I_start: initial number of COVID-19 cases       
*R_start: initial number of recovered

@@

The parameter $\sigma$ is our parameter of interest; we estimate $\sigma$ fixing $E_{start}= 3*I_{start}$ minimizing the log likelihood of the described Poisson distribution (or Negative Binomial one).

@@

f1<-function(par){
sigma<-expit(par)
f_exp<-3*I_1/N
parameters <- c(mu = mu0, beta =(mean(beta_vec[1:(days-last-2)])+1/duration), sigma = sigma, gamma =1/duration)
initials <- c(S = (1-(f_exp+I_1/N-R_start/N)), E = f_exp, I = I_1/N, R = R_1/N)
-LogLikfun(initials,parameters,dat_csv$totale_attualmente_positivi[c((days-last):days)])
}
est<-optim(fn=f1,par=logit(0.1))
#estimated par
expit(est$par)
sigma<-expit(est$par)
f_exp<-3*I_1/N
parameters <- c(mu = mu0, beta =(mean(beta_vec[1:(days-last-2)])+1/duration), sigma = sigma, gamma = 1/duration)
initials <- c(S = (1-(f_exp+I_1/N-R_1/N)), E = f_exp, I = I_1/N, R = R_1/N)
pro<-SEIR(pars = parameters, init = initials, time = 0:last)
#predicted vs observed
cbind(pro$results$I*N,dat_csv$totale_attualmente_positivi[c((days-last):days)])
```

@@

```{r status, echo=TRUE}
#the unknowrn fraction of exposed people is
pro$results$E[last+1];
#that is 
pro$results$E[last+1]/pro$results$I[last+1]
#times the infected
```
For the beta parameter, we perfom a simulation on its trend by means of a Bayesian Structural Time Series using the library bsts of R.    
We estimate a BSTS model specifing a local linear trend ans 1000 simulation.

### 2020-03-24 draft_analysis_Veneto_html.Rmd
Add home page, navigation and theme

We made a forecast by means of a SEIR model fixing a series of initial status:
*S:N, the size of the Veneto Region population  
*E: The number of exposed people (unknown)  
*I_start: initial number of COVID-19 cases     
*R_start: initial number of recovered

and parameters:  
- beta: gamma + slope
- gamma= 1/duration of diseases (duration=21 days)
- sigma: the coronavirus transmission rate 
- mu0: the overall mortality rate

@@

f1<-function(par){
f_exp<-expit(par[1])
#max 0.33 according to https://www.nejm.org/doi/10.1056/NEJMoa2001316 is 1/5.2
sigma<-expit(par[2])/3
parameters <- c(mu = mu0, beta =(as.numeric(beta_vec[days-last-2])+1/duration), sigma = sigma, gamma =1/duration)
initials <- c(S = (1-(f_exp+I_1/N+R_start/N)), E = f_exp, I = I_1/N, R = R_1/N)
-LogLikfun(initials,parameters,dat_csv$totale_attualmente_positivi[c((days-last):days)])
}
est<-optim(fn=f1,par=c(logit(0.02),logit(0.2)))
#estimated par
expit(est$par)

f_exp<-expit(est$par[1])
sigma<-expit(est$par[2])/3
parameters <- c(mu = mu0, beta =(as.numeric(beta_vec[days-last-2])+1/duration), sigma = sigma, gamma = 1/duration)
initials <- c(S = (1-(f_exp+I_1/N+R_1/N)), E = f_exp, I = I_1/N, R = R_1/N)
pro<-SEIR(pars = parameters, init = initials, time = 0:last)
#predicted vs observed
cbind(pro$results$I*N,dat_csv$totale_attualmente_positivi[c((days-last):days)])
```

@@

```{r status, echo=TRUE}
#mthe unknowrn fraction of exposed people is
pro$results$E[last+1]
#that is 
pro$results$E[last+1]/(I_1/N)
#times the infected
```

@@

```{r bsts model fit and prediction, echo=TRUE}
# Bayesian Structural Time Series
#we drop the first 6 beta_vec
ss <- AddLocalLinearTrend(list(), beta_vec[6:(days-4)])
ss <- AddSeasonal(ss, beta_vec[6:(days-4)], nseasons = 7)
model1 <- bsts(beta_vec[6:(days-4)],
               state.specification = ss,
               niter = 1000,seed=123)
par(mfrow = c(1,1))
plot(model1, "components")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Beta1 coefficient",main="Data and predictions",ylim=c(-1,1)*IQR(pred1$distribution)+median(pred1$distribution))
# matrix of beta coefficients
coef<-(pred1$distribution[,3:16])
par(mfrow = c(1,1))
```

@@

```{r  scenario2 sim and plot , echo=TRUE}
ss<-NULL
set.seed(123)
ss <- AddLocalLinearTrend(list(), rnorm(100,beta_vec[days-4],sd_vec[days-4]))
model1 <- bsts(beta_vec[6:(days-4)],
               state.specification = ss,
               niter = 1000,seed=123)
par(mfrow = c(1,1))
plot(model1, "components")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Beta1 coefficient",main="Data and predictions",ylim=c(-1,1)*IQR(pred1$distribution)+median(pred1$distribution))
# matrix of beta coefficients
coef<-(pred1$distribution[,3:16])
par(mfrow = c(1,1))

@@

### confidence limits
I_seir_med<-tapply(seir2_sim$I,seir2_sim$time,median)
I_seir_lwr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.25)
I_seir_upr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.75)

days.before<-date[1:days]
days.ahead<-date[(days+1):(days+forecast)]
step.ahead<-forecast+1
mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Veneto Region: Scenario 2 (Credible Interval ",100*0.50,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p
```

@@

I_seir_med<-tapply(seir1_sim$I,seir1_sim$time,median)
I_seir_lwr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.25)
I_seir_upr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.75)

>>>>

We made a forecast by means of a SEIR model fixing a series of initial status:  
*S:N, the size of the Veneto Region population  
*E: The number of exposed people, but supposed to be at least 3 times the infected (3 x I_start)  
*I_start: initial number of COVID-19 cases     
*R_start: initial number of recovered

and parameters:   
-beta: gamma + slope   
-gamma= 1/duration of diseases (duration=21 days)   
-sigma: the coronavirus transmission rate   
-mu0: the overall mortality rate

@@

f1<-function(par){
#max 0.20 according to https://www.nejm.org/doi/10.1056/NEJMoa2001316 is 1/5.2
sigma<-expit(par)
f_exp<-3*I_1/N
parameters <- c(mu = mu0, beta =(as.numeric(beta_vec[(days-last-2)])+1/duration), sigma = sigma, gamma =1/duration)
initials <- c(S = (1-(f_exp+I_1/N+R_start/N)), E = f_exp, I = I_1/N, R = R_1/N)
-LogLikfun(initials,parameters,dat_csv$totale_attualmente_positivi[c((days-last):days)])
}
est<-optim(fn=f1,par=logit(0.1))
est
expit(est$par)

f_exp<-3*I_1/N
sigma<-expit(est$par)

parameters <- c(mu = mu0, beta =(as.numeric(beta_vec[(days-last-2)])+1/duration), sigma = sigma, gamma = 1/duration)
initials <- c(S = (1-(f_exp+I_1/N+R_1/N)), E = f_exp, I = I_1/N, R = R_1/N)
pro<-SEIR(pars = parameters, init = initials, time = 0:last)
#predicted vs observed
cbind(pro$results$I*N,dat_csv$totale_attualmente_positivi[c((days-last):days)])
```

@@

```{r status, echo=TRUE}
#mthe unknowrn fraction of exposed people is
pro$results$E[last+1]
#that is 
pro$results$E[last+1]/pro$results$I[last+1]
#times the infected
```

@@

```{r bsts model fit and prediction, echo=TRUE}
# Bayesian Structural Time Series
#we drop the first 6 beta_vec
ss <- AddStudentLocalLinearTrend(list(), beta_vec[6:(days-4)])
ss <- AddSeasonal(ss, beta_vec[6:(days-4)], nseasons = 7)
model1 <- bsts(beta_vec[6:(days-4)],
               state.specification = ss,
               niter = 1000,seed=123)
par(mfrow = c(1,1))
plot(model1, "components")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Beta1 coefficient",main="Data and predictions",ylim=c(-1,1)*IQR(pred1$distribution)+median(pred1$distribution))
# matrix of beta coefficients
coef<-(pred1$distribution[,3:16])
par(mfrow = c(1,1))
```

@@

```{r  scenario2 sim and plot , echo=TRUE}
ss<-NULL
set.seed(123)
ss <-   AddLocalLevel(list(), beta_vec[(days-9):(days-4)],sigma=SdPrior(sd_vec[(days-4)]))
model1 <- bsts(beta_vec[(days-9):(days-4)],
               state.specification = ss,
               niter = 1000,seed=123)
par(mfrow = c(1,1))
plot(model1, "components")
#previsioni
pred1 <- predict(model1, horizon = 16, burn = 100)
par(mfrow = c(1,1))
plot(pred1 , ylab="Beta1 coefficient",main="Data and predictions",ylim=c(-1,1)*IQR(pred1$distribution)+median(pred1$distribution))
# matrix of beta coefficients
coef<-(pred1$distribution[,3:16])
par(mfrow = c(1,1))

@@

### confidence limits
I_seir_med<-tapply(seir2_sim$I,seir2_sim$time,median)
I_seir_lwr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.05)
I_seir_upr<-tapply(seir2_sim$I,seir2_sim$time,quantile,p=0.95)

days.before<-date[1:days]
days.ahead<-date[(days+1):(days+forecast)]
step.ahead<-forecast+1
mu.lower<-c(dat_csv$totale_attualmente_positivi,I_seir_lwr*N)
mu.upper<-c(dat_csv$totale_attualmente_positivi,I_seir_upr*N)
mu.med<-xts(c(dat_csv$totale_attualmente_positivi,I_seir_med*N),order.by = c(days.before,days.ahead),frequency = 7)
counts<-mu.med
mu<-xts(x = as.matrix(cbind(counts,mu.lower,mu.upper)) , order.by = c(days.before,days.ahead))
p <- dygraph(mu,main=paste("Veneto Region: Scenario 2 (Credible Interval ",100*0.90,"%)",sep = ""),ylab=" Infected",xlab="Day",height=400,width=800) %>%  dySeries(c("mu.lower", "counts", "mu.upper"),label="counts")
p<-p %>% dyLegend(show = "always", hideOnMouseOut = FALSE) %>%  dyShading(from = days.ahead[1], to = days.ahead[step.ahead], color = "#CCEBD6")%>% dyEvent(days.ahead[1], "Prediction", labelLoc = "bottom")
p
```

@@

I_seir_med<-tapply(seir1_sim$I,seir1_sim$time,median)
I_seir_lwr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.05)
I_seir_upr<-tapply(seir1_sim$I,seir1_sim$time,quantile,p=0.75)
